---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# imfeatures



<!-- badges: start -->
[![R-CMD-check](https://github.com/bbuchsbaum/imfeatures/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbuchsbaum/imfeatures/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`imfeatures` extracts visual features from 2D images using deep learning
models and traditional computer vision techniques. It supports VGG16,
ResNet, and other pre-trained models via Python (Keras / `thingsvision` /
CLIP), and provides methods for edge entropy, multiscale entropy, and
feature-similarity metrics.

## Install

```r
# install.packages("devtools")
devtools::install_github("bbuchsbaum/imfeatures")
```

The package depends on Python for deep-learning backends. On a workstation,
`imfeatures_config()` sets up a managed environment automatically; on HPC
or other shared systems, see the HPC Quickstart below.

## Cookbook

Compute pairwise feature similarities at three VGG-16 layers:

```{r example, eval = FALSE}
library(imfeatures)
imgs <- list.files(system.file("extdata", package = "imfeatures"),
                   pattern = "\\.(jpe?g|png)$", full.names = TRUE)
im_feature_sim(imgs, layers = c(1, 2, 3))
```

Compute edge entropy on a grayscale matrix (no Python required):

```{r entropy, eval = FALSE}
library(imfeatures)
img <- matrix(runif(100 * 100), nrow = 100)
compute_edge_entropy(img)
```

## HPC Quickstart (single recommended path)

On HPC systems, avoid automatic Conda setup and use an existing Python instead.

1. Disable auto Python setup during load (recommended on HPC):

```r
Sys.setenv(IMFEATURES_SKIP_PYTHON = "TRUE")
```

To make this permanent, add `IMFEATURES_SKIP_PYTHON=TRUE` to `~/.Renviron`.

2. Create or choose a Python environment (module + venv is typical):

```bash
# Example using your cluster's Python module
module load python/3.10             # if applicable (3.9/3.10 recommended)
python -m venv $WORK/venvs/imfeatures
source $WORK/venvs/imfeatures/bin/activate
pip install --upgrade pip wheel setuptools

# Minimal required packages
pip install Pillow numpy

# Optional packages for full functionality
pip install thingsvision resmem open-clip-torch

# PyTorch: use your cluster's module if available, otherwise pick one index
# CPU only:
pip install torch torchvision --index-url https://download.pytorch.org/whl/cpu
# or CUDA-specific (example):
# pip install torch torchvision --index-url https://download.pytorch.org/whl/cu118
 
Note on Python versions: some optional packages (e.g., thingsvision via numba)
currently do not support Python 3.11+. Prefer Python 3.9 or 3.10 when you plan
to use these features.
```

3. Tell imfeatures to use that Python:

```r
library(imfeatures)
use_existing_python(file.path(Sys.getenv("WORK"), "venvs/imfeatures/bin/python"))
```

Alternatively, set once in `~/.Renviron` and it will be auto-detected:

```
RETICULATE_PYTHON=/full/path/to/venvs/imfeatures/bin/python
IMFEATURES_SKIP_PYTHON=TRUE
```

4. Verify:

```r
reticulate::py_config()
```

Troubleshooting:

- If you see a Conda error like “bad interpreter” during `library(imfeatures)`, it’s usually a broken R-miniconda on shared filesystems. Use the HPC Quickstart above or set `IMFEATURES_SKIP_PYTHON=TRUE` and call `use_existing_python()`.
- To force `virtualenv` instead of Conda during `imfeatures_config()`, set `IMFEATURES_METHOD=virtualenv` before calling it.
