---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# imfeatures



<!-- badges: start -->
[![R-CMD-check](https://github.com/bbuchsbaum/imfeatures/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbuchsbaum/imfeatures/actions/workflows/R-CMD-check.yaml)
[![Codecov coverage](https://codecov.io/gh/bbuchsbaum/imfeatures/branch/main/graph/badge.svg)](https://app.codecov.io/gh/bbuchsbaum/imfeatures)
<!-- badges: end -->

The goal of imfeatures is to ...

## Get started

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bbuchsbaum/imfeatures")
```

## Cookbook

This is a basic example which shows you how to solve a common problem:

```{r example}
library(imfeatures)
## basic example code
im1 = "testdata/1_A_1.jpeg"
im2 = "testdata/1_B_1.jpeg"

im_feature_sim(c(im1,im2), layers=c(1,2,3))
```

## HPC Quickstart (single recommended path)

On HPC systems, avoid automatic Conda setup and use an existing Python instead.

1. Disable auto Python setup during load (recommended on HPC):

```r
Sys.setenv(IMFEATURES_SKIP_PYTHON = "TRUE")
```

To make this permanent, add `IMFEATURES_SKIP_PYTHON=TRUE` to `~/.Renviron`.

2. Create or choose a Python environment (module + venv is typical):

```bash
# Example using your cluster's Python module
module load python/3.9              # if applicable
python -m venv $WORK/venvs/imfeatures
source $WORK/venvs/imfeatures/bin/activate
pip install --upgrade pip wheel setuptools

# Minimal required packages
pip install Pillow numpy

# Optional packages for full functionality
pip install thingsvision resmem open-clip-torch

# PyTorch: use your cluster's module if available, otherwise pick one index
# CPU only:
pip install torch torchvision --index-url https://download.pytorch.org/whl/cpu
# or CUDA-specific (example):
# pip install torch torchvision --index-url https://download.pytorch.org/whl/cu118
```

3. Tell imfeatures to use that Python:

```r
library(imfeatures)
use_existing_python("$WORK/venvs/imfeatures/bin/python")
```

Alternatively, set once in `~/.Renviron` and it will be auto-detected:

```
RETICULATE_PYTHON=$WORK/venvs/imfeatures/bin/python
IMFEATURES_SKIP_PYTHON=TRUE
```

4. Verify:

```r
reticulate::py_config()
```

Troubleshooting:

- If you see a Conda error like “bad interpreter” during `library(imfeatures)`, it’s usually a broken R-miniconda on shared filesystems. Use the HPC Quickstart above or set `IMFEATURES_SKIP_PYTHON=TRUE` and call `use_existing_python()`.
- To force `virtualenv` instead of Conda during `imfeatures_config()`, set `IMFEATURES_METHOD=virtualenv` before calling it.

