---
title: "Build ordered image-feature hierarchies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build ordered image-feature hierarchies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

When several network stages describe the same images, the first principal
components of every stage often recover the same high-variance trajectory.
A hierarchy transform turns those aligned blocks into compact scores and
then into **innovations**: information that first becomes linearly available
at a given stage.

You start with a named list of matrices — one stage per block, same rows —
and finish with a named list you can pass to an encoding model.

```{r load}
library(imfeatures)
```

## What do the inputs look like?

The vignette uses a synthetic three-stage movie: a shared slow trajectory
plus stage-specific signal. In a real analysis the blocks would be VGG
activations after spatial pooling and, if you are modelling fMRI, after
HRF convolution.

```{r simulate-blocks}
set.seed(1)
n <- 120
t <- seq_len(n) / n
shared <- cbind(sin(2 * pi * t), cos(2 * pi * t))

stage <- function(scale, p, noise) {
  unique <- matrix(rnorm(n * 2), n, 2)
  cbind(scale * shared, unique) %*% matrix(rnorm(4 * p), 4, p) +
    matrix(rnorm(n * p, sd = noise), n, p)
}

blocks <- list(
  b1 = stage(1.0, 18, 0.15),
  b2 = stage(0.8, 18, 0.15),
  b3 = stage(0.5, 18, 0.15)
)
```

```{r check-blocks, include = FALSE}
stopifnot(length(unique(vapply(blocks, nrow, integer(1)))) == 1L)
stopifnot(all(vapply(blocks, function(m) all(is.finite(m)), logical(1))))
```

## How do you compile the hierarchy?

Write a recipe, fit it on training rows, then apply it. The fit stores
PCA maps and cross-stage coefficients, so each new row is transformed
from its own stage values only.

```{r fit-recipe}
recipe <- hierarchy_recipe(
  order = c("b1", "b2", "b3"),
  compress = block_pca(rank = 8, score_weighting = "whitened"),
  decomposition = ordered_innovation(method = "svd"),
  name_prefix = "vgg"
)
fit <- fit_hierarchy_transform(recipe, blocks)
features <- predict(fit, blocks)
```

```{r show-names}
names(features)
```

```{r check-fit, include = FALSE}
stopifnot(identical(names(features), c("vgg_b1", "vgg_b2_innovation", "vgg_b3_innovation")))
stopifnot(all(vapply(features, ncol, integer(1)) == 8L))
stopifnot(all(vapply(features, function(m) all(is.finite(m)), logical(1))))
```

The first band keeps the earliest stage. Later bands are innovations after
removing everything already linearly available from earlier stages.

## Did the decomposition separate the stages?

On the rows used to fit the maps, later innovations should be orthogonal
to earlier scores. Linear CKA between those innovation blocks should be
near zero.

```{r diagnostics}
diag <- hierarchy_diagnostics(fit, blocks)
diag$innovation[, c("stage", "residual_energy", "fraction_predictable")]
```

```{r show-cka}
round(diag$pairwise_cka, 3)
```

```{r check-diag, include = FALSE}
stopifnot(abs(diag$pairwise_cka["b1", "b2"]) < 1e-8)
stopifnot(abs(diag$pairwise_cka["b1", "b3"]) < 1e-8)
stopifnot(all(diag$innovation$residual_energy >= 0, na.rm = TRUE))
stopifnot(all(diag$innovation$residual_energy <= 1 + 1e-8, na.rm = TRUE))
stopifnot(all(is.finite(as.matrix(diag$compression[, c("cka_retained", "recon_error")]))))
```

Residual energy is the fraction of each compressed stage that remains after
the earlier stages are projected out. A small value means that stage mostly
repeated earlier information.

```{r residual-energy-plot, echo = FALSE, fig.width = 6, fig.height = 3.2, fig.alt = "Bar plot of residual energy by stage"}
op <- par(mar = c(4, 4, 1, 1))
barplot(
  diag$innovation$residual_energy,
  names.arg = diag$innovation$stage,
  ylim = c(0, 1),
  ylab = "Residual energy",
  col = "grey80",
  border = NA
)
par(op)
```

## Can you apply the same maps to new rows?

Split the simulated movie, fit on the first half, and transform both halves.
Batch membership cannot change a row: transforming 60 test rows together
must match transforming any subset of those rows.

```{r held-out}
train <- subset_feature_blocks(blocks, 1:60)
test <- subset_feature_blocks(blocks, 61:120)
fit_half <- fit_hierarchy_transform(recipe, train)
test_features <- predict(fit_half, test)
invariance <- check_transform_invariance(fit_half, test)
```

```{r show-invariance}
invariance[c("batch_invariant", "permutation_equivariant")]
```

```{r check-heldout, include = FALSE}
stopifnot(isTRUE(invariance$batch_invariant))
stopifnot(isTRUE(invariance$permutation_equivariant))
held <- hierarchy_diagnostics(fit_half, train, heldout = test)
stopifnot(is.finite(held$heldout$pairwise_cka["b1", "b2"]))
```

Held-out CKA is not forced to zero. If it rebounds toward 1, the
decomposition overfit the training segment.

```{r heldout-cka}
held <- hierarchy_diagnostics(fit_half, train, heldout = test)
round(held$heldout$pairwise_cka, 3)
```

## Where do image models enter?

`vgg16_block_end_spec()` records one tap at the end of each VGG-16 block
and a depth-adaptive DCT pooler. `extract_feature_hierarchy()` returns the
named raw blocks; you then fit the same recipe. Temporal alignment such as
HRF convolution stays outside this package.

```{r spec-example, eval = FALSE}
spec <- vgg16_block_end_spec()
raw <- extract_feature_hierarchy(image_paths, spec = spec)
fit <- fit_hierarchy_transform(recipe, raw)
```

`residualize_tiers()` remains a compatibility wrapper around the same
inductive maps.

## Next steps

- `?hierarchy_recipe` and `?fit_hierarchy_transform` for compression and
  innovation options
- `?vgg16_block_end_spec` for the default VGG-16 taps
- `?hierarchy_diagnostics` for stimulus-only checks before any neural data
